---
title: "rankMotifs_groups.Rmd"
author: "Thomas Manke"
date: "`r date() `"
output:
  
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
    toc_float: yes
    code_folding: hide
  html_notebook:
    
    code_folding: hide
    df_print: paged
    fig_height: 10
    fig_width: 10
    toc: yes
    toc_depth: 2
    toc_float: yes

# set default files (examples) and directory
params:
  f1: "trap.Rds"
  f2: "liver.tsv"
  pid: !r Sys.getpid()
  dn: "rankMotifs_groups"
  reduceA: FALSE
   
# runtime: shiny  # uncomment if shiny part is to be run below
---
# Intro
This is a notbook version of a simple R-script
https://github.com/maxplanck-ie/TRAP/blob/master/utilities/rankMotifs_groups.R

Input:

1. a sequence-motif matrix with TRAP-scores
2. a list of sequence identifiers (optionally with group labels)

Output: Ranked list of motifs according to their discriminative power for the sequences

# Setup
```{r setup}
# set global knitr options
knitr::opts_chunk$set(warning = FALSE, eval=TRUE)
knitr::opts_chunk$set(root.dir=getwd())             # default: location of the markdown file

# load libraries
library(parallel)
library(tidyverse)

# define functions
# passing sequence ids (seq1, seq2) explicitly
#wt = function(x) { wilcox.test (A[seq1,x], A[seq2,x])$p.value  }   # two-sided wilcox p-value
#lf = function(x) { log2(  mean(A[seq1,x])/mean(A[seq2,x])  )   }   # log2-fold change

# passing group-ids
lf  = function(x) { rr=aggregate(A[,x] ~ groups, as.data.frame(A[,x]), mean ); return (log2(rr[1,2]/rr[2,2])) }
wt  = function(x) { wilcox.test (A[,x] ~ groups)$p.value  }         # two-sided wilcox p-value
kt  = function(x) { kruskal.test(A[,x] ~ groups)$p.value }          # kruskal test for multiple groups  
lmf = function(x) { lm( A[,x] ~ groups )  }                         # define linear model of TRAP-score as function of group


# define global variables
myred=rgb(1,0,0,0.5)
myblue=rgb(0,0,1,0.5)
myyellow=rgb(1,1,0,0.5)
```


# Read Arguments
```{r read_arguments}
usage=
"This script takes as input
	1. a sequence-motif matrix A of TRAP-scores: trap.mat, (Rds file)
	2. a list seqs of sequence ids (optionally with group labels): gene_list.tsv (ascii [2-column]) 
	3. an output directory name dn:  will be recreated (recursively)
	4. a flag \"reduceA\" to indicate whether the matrix A should be reduced to entries listed in seqs (default: FALSE)

The script/markdown outputs a directory with *tsv files of ranked motif list according to some group test
        a. For two groups a Wilcoxon test is performed (wilcox.tsv)
	      b. For two or more groups a Kruskal-Wallis test is performed (kruskal.tsv)
        c. A box plot for the top motifs is also generated (kruskal.jpg)
        d. a markdown report (html)
Usage: R -e \"rmarkdown::render('rankMotifs_groups.Rmd',  params = list(f1 = \"file.Rds\", f2 = \"list.tsv\", dn=\"output_dir\" ) )\" 

\n"


# Copy-Paste Example: 
# Rscript -e "rmarkdown::render('~/R/Rstudio/TRAP/rankMotifs_groups.Rmd', params = list(dn='test')) "
warning(usage)

pid=params$pid
f1=params$f1
f2=params$f2
dn=paste0(params$dn,".", pid)
reduceA=params$reduceA

dir.create(dn, recursive=TRUE)
cat("write results to directory", dn, "\n")
```

# Read Data
```{r read_data}
A    = readRDS(f1)                   # read matrix of TRAP-scores

cat("Read matrix A: ", dim(A), " from file ", f1, "\n")
ir=min(3,nrow(A))
ic=min(3,ncol(A))
A[1:ir,1:ic]

seqs = read.table(f2,row.names=1)    # read list of sequence. Possibly with unique group annotations (seqs$V2). No duplicated rownames
cat("Read seqs: ", dim(seqs), " from file ", f2, "\n")
head(seqs,3)
```

# Filtering

```{r filter}
if (reduceA) {
  i=which(rownames(seqs) %in% rownames(A))
  A=A[i,]
}

N = nrow(A)         # number of sequences
M = ncol(A)         # number of motif matrices
nlist=rownames(A)   # list of all sequence_ids
mlist=colnames(A)   # list of all matrix_ids

cat("Dimension of A after filtering:", dim(A), "\n")
```

# Groups
```{r groups}
# assign unique group names (=factors) to sequences
if (ncol(seqs)==0) { seqs$V2=factor("signal") }    # if groups are not explicitly named in seqs, then call this group "signal"
groups=gl(1, nrow(A))                              # create groups for all genes in A
levels(groups)=c("backgroundA",levels(seqs$V2))    # define levels of all groups in A, default group: "backgroundA"
i=which(rownames(A) %in% rownames(seqs))
groups[i]=seqs[rownames(A[i,]), "V2"]              # set group level to those defined by seqs

#table(groups)                                     # some levels may not have any support if they are filtered out
groups=droplevels(groups)                          # drop all levels that are not supported by data in A (i.e. groups that only exist in list seqs)
nl=nlevels(groups)                                 # number of levels defined in groups 
tb=table(groups)                                      # should not have any zero entries
```


```{r show_table}
cat(N," sequences are divided in ", nl, " groups from file ", f2, "\n")
#groups %>% as_data_frame() %>% ggplot(aes(x=groups)) + geom_bar() + geom_text(stat='count', aes(label=..count..), color="red", vjust=-0.5)

groups %>% as_data_frame() %>% ggplot(aes(x=groups)) + geom_bar() + geom_text(stat='count', aes(label=..count..), color="red") + coord_flip()

tb
```


########################################################################
```{r cluster, eval=FALSE, echo=FALSE}
# Initialize compute cluster for parallel processing: --> parSapply
cl=makeCluster(2)
clusterExport(cl,c("groups","A"))   # pass objects to cluster
```

# Wilcoxon Test
If there are only two sets, define the motifs with significant TRAP differences between the two sets.
```{r wilcox}
if (nl == 2) {
  
	log2FC<-sapply(mlist, lf)     # fold-change calculation over list of motifs mlist
	pvalues<-sapply(mlist, wt)    # wilcoxon test pvalue calculation over list of motifs mlist

	#log2FC<-parSapply(cl, mlist, lf)     # parallelize fold-change calculation over list of motifs mlist
	#pvalues<-parSapply(cl, mlist, wt)    # parallelize wilcoxon test pvalue calculation over list of motifs mlist
	fdr = p.adjust(pvalues, method="fdr")

	res = rbind(log2FC = log2FC, pvalue=pvalues, fdr=fdr )
	res = t(res)                            # transpose
	res = res[order(res[,"pvalue"]),]       # order by p-values

	fn=file.path(dn,"wilcox.tsv")
	cat(file=fn, "#Wilcoxon test. N=",N," M=",M,"\n")
	cat(file=fn, "#groups: ",names(tb), "\n", append=TRUE)
	cat(file=fn, "#sizes:  ",tb,"\n", append=TRUE)
	cat(file=fn, "#matrix","log2FC", "pvalue","fdr","\n",sep="\t",append=TRUE)
	write.table(file=fn, res, quote=FALSE,col.names=FALSE, sep="\t", append=TRUE)
  cat("Wrote results to ", fn, "\n")
}
```

########################################################################
# Kruskal Test 
The Kruskal test extends Wilcoxon to multiple groups of sequences (includes nl=2)
```{r kruskal}
#pvalues<-parSapply(cl,mlist,kt)      # parallelize Kruskal test for each motif in mlist
pvalues<-sapply(mlist,kt)             # Kruskal test for each motif in mlist
fdr=p.adjust(pvalues,method="fdr")
res=data.frame(pvalue=pvalues,fdr=fdr)

res=res[order(res[,"pvalue"]),]                                              # sort results by pvalue
fn=paste(dn,"/kruskal.tsv",sep="")                                           # filename for output
cat(file=fn,"#Kruskal test. N=", N, " M=", M, "\n")                          # print header information
cat(file=fn,"#groups:", levels(groups), "\n",append=TRUE)
cat(file=fn,"#sizes: ", table(groups), "\n",append=TRUE)
cat(file=fn,"#matrix", names(res), "\n", sep="\t", append=TRUE)
write.table(file=fn, res,quote=FALSE, col.names = FALSE, sep="\t",append=TRUE)

cat("Wrote results to ", fn, "\n")

# Kruskal generalizes Anova
# Anova model: normality assumption
#models       = parSapply(cl,  mlist, lmf, simplify=FALSE)
#anova_tables = parSapply(cl, models, anova, simplify=FALSE)
#anovap=lapply(anova_tables, function(x) x$"Pr(>F)"[[1]])
#res$anovap=unlist(anovap)
```




# Plotting
```{r plots}
# define top-significant motifs
fdr_thresh=0.05
top=which(res[,"fdr"]<fdr_thresh)          # define top results (based on Kruskal-test (FDR)
if (length(top)<3) { top=1:3 }             # show at least the top three motifs

# prepare for A for efficient boxplot
DF=as.data.frame(A[,rownames(res)[top]])   # convert top entries of A to data frame: 
mn=names(DF)                               # keep motif names
DF$groups=groups                           # assign groups to DF

# library(reshape2)
# DF=melt(DF, value.name="TRAP", id.vars="groups", variable.name="motif")  # old: reshape2::melt

DF = DF %>% gather(key="motif", value="TRAP", -groups) %>% mutate(motif = as.factor(motif))
DF = DF %>% mutate(motif=sprintf("%s \n %2.3e",motif, res[motif,"fdr"]))      # add (rounded) fdr as new line to motif name

gp = ggplot(DF, aes(x=motif,y=TRAP, col=groups)) + geom_violin() + coord_flip()   # ggplot


gp

 # send plot to jpeg file
fn=file.path(dn,"kruskal.jpg") 
jpeg(fn)                                                                    
gp
do=dev.off()

cat("Sent boxplot to ", fn, "\n")
```


```{r old_boxplot, eval=FALSE}
# similar with base R
fn=file.path(dn,"kruskal.old.jpg") 
jpeg(fn)                                                                    
bp=boxplot(TRAP~groups*motif,data=DF,col=rainbow(nl),horizontal=TRUE,las=1,boxwex=0.5,xlab="TRAP-score",notch=TRUE, main="Significant Motifs",yaxt="n")
xp=min(DF$TRAP)*0.95                     # x-position  for text
yps=0.5 + seq(1, nl*length(mn), by=nl)   # y-positions for text 
fdrs=sprintf("%2.3e",res[top,"fdr"])     # fdr-values for each motif to be plotted at (x,y)
axis(2, labels=mn,at=yps,las=2)                                                 # plot y-axis with motif names
text(xp, yps,labels = fdrs,cex = 0.75)                                          # set text (fdr-values)
legend("topright",inset=0.03, legend=levels(groups), fill=rainbow(nl), cex=0.8) # plot legend for nl group levels
do=dev.off()

cat("Sent boxplot to ", fn, "\n")
```


```{r shiny, eval=FALSE, echo=FALSE}
# shiny renders the dynamic ggplot poorly. None of the following suggestions fixed the problem
#1. return plot-Object (ggp) rather than plotting witnin renderPlot
#2. use different library  
# library(Cairo)
# options(shiny.usecairo=T)
# 3. safe intermediate file and  renderImage() --> https://shiny.rstudio.com/articles/images.html (not yet tried)

numericInput("n", "Number of top motifs", 3, min=2)

renderPlot({
  top=1:input$n
  DF=as.data.frame(A[,rownames(res)[top]])             # convert top entries of A to data frame: 
  # the following is safer (for top=1:1), but slower (first convert then select)
  # instead enforce input$n >= 2
  # DF=as.data.frame(A) %>% select(rownames(res)[top]);  
  
  mn=names(DF)                               # keep motif names
  DF$groups=groups                           # assign groups to DF
  DF = DF %>% gather(key="motif", value="TRAP", -groups) %>% mutate(motif = as.factor(motif))
  DF = DF %>% mutate(motif=sprintf("%s \n %2.3e",motif, res[motif,"fdr"]))         # add (rounded) fdr as new line to motif name
  ggp = ggplot(DF, aes(x=motif,y=TRAP, col=groups)) + geom_violin() + coord_flip()   # create ggplot object
  ggp # return ggplot object       
})
```

```{r final}
#stopCluster(cl)
sessionInfo()

# deprecated: saving all data objects for subsequent analysis
# this was never much used and may leave an unnecessarily large storage footprint
# sn=paste0(dn, "/rankMotifs_groups.Rdata")
# save.image(file=sn)
```
